\Large\textbf{3D Graphics Rendering with OpenGL}

### 2019-04-03 Paper.tex
Revise

% Title Page
\begin{titlingpage}
    \maketitle
    \begin{abstract}
    Computer graphics is an essential component to any consumer facing computer. Efficiently rendering computer graphics requires the use of specialized hardware in the form of graphics processing units. GPUs work differently compared to the CPUs which programmers are experienced with due to the former's focus on parallelization. Graphics APIs have been created to help programmers write performant and portable code for GPUs to render computer graphics. This paper will introduce the core concepts of 3D graphics rendering and OpenGL, a popular and widely-supported graphics API.
    \end{abstract}
\end{titlingpage}

% Body
\section{Introduction}
Computers have become widespread and have touched many aspects of modern-day life. From new forms of entertainment such as video games to artificial intelligence they have revolutionized the way to work, learn, and play. Computer graphics in particular have been a driving force behind the adoption of computers because they allow users to easily interact with computers. Programmers use computer graphics to display user interfaces, render video games, and even animate entire films. Every day millions of people interact with interfaces on phones, laptops, and other devices that allow them to focus on the work they are doing rather than how they communicate with their device. % TODO maybe this could still be improved

Plenty of tools to manipulate 2D and 3D graphics already exist, such as: 3D object editors, image processing applications, UI libraries, and game engines. These are effective tools to quickly get work done, but the user may not know how their work is being accomplished. Without this understanding it may be harder for the programmer to debug or optimize their code, and by learning the lower-level concepts of computer graphics, one is able to better use higher-level abstractions and tools. % TODO this is still a little awkward

This paper intends to introduce the reader to the broad concepts of graphics processing. This includes the purpose of graphics processing units, why graphics APIs exist, and an introduction to a commonly used cross-platform graphics API --- OpenGL. OpenGL will be discussed so that the reader has an overview of the concepts of OpenGL, and can create an OpenGL program. This paper will focus on the use OpenGL version 3.2 which is widely supported and is very similar to updated versions of OpenGL.

\section{Background}
Graphics processing is predisposed towards parallel processing due to the repetitive and independent nature of the calculations that it requires \cite[p.~4]{sellers2016}. Graphics processing units (GPUs) were created in order to meet the unique hardware requirements of computer graphics, and are fundamentally different from more traditional central processing units (CPUs). While CPUs generally have a few very fast cores, GPUs take an alternate approach of having a much greater number of slower cores. Figure \ref{fig:architecture} illustrates this difference in architecture, showing the massive difference in arithmetic logic units (ALUs) between CPUs and GPUs. These ALUs give GPUs an edge over CPUs when performing mathematical operations and are a core part of the parallel processing capability that GPUs possess.

@@

A difference also exists between the two in their approach to threading. An application running on a CPU may generally have only few threads running due to the high cost of creation and context switches. GPUs have lower cost thread creation and can more quickly context switch, and they often require tens of thousands of threads in order to fully reach their processing potential \cite{larkin2016}. Despite the slower performance of an individual GPU core, GPUs easily outperform CPUs in tasks related to graphics processing due this difference in architecture which gives them an edge in parallel processing. Figure \ref{fig:performance} shows the historical theoretical performance gap between CPUs and GPUs in giga floating-point operations per second (GFLOP/s), a common measure of graphical computation speed. The graph shows that GPUs have clear performance advantage. Using GPUs effectively is especially important when rendering graphics since they often render items such as user interfaces which generally should remain responsive at all times.

@@

Because of the architectural differences between GPUs and CPUs, GPUs are programmed differently than CPUs. In addition to writing high-level code which is compiled into assembly, GPUs are also programmed by using APIs that are exposed by the manufacturer of the graphics card. GPU manufacturers may support proprietary APIs which is very common on cards that are made to be used in systems such as game consoles and arcade machines, or they may implement standardized APIs such as OpenGL, Metal, Vulkan, or DirectX.

While using a proprietary API may allow a programmer to get more performance out of the hardware they are targeting, it will severely limit the application's portability. A program that is written to run on a GPU using OpenGL should be compatible with any other GPU implementing the same version of OpenGL or higher. These APIs help programmers to focus on the code they are writing rather than what hardware it will be executed on and are created so that they are at a low enough level that maximum performance can be achieved while also being at a high enough level so that programmers can easily use them \cite[p.~4]{sellers2016}.

\section{Graphics Rendering}
Computer displays such as monitors use a 2D matrix of pixels to display images. The dimensions of this matrix is referred to as the display's resolution. Traditionally each pixel of a color display has a red, blue, and green value which all together determine the color of the pixel. Graphics rendering is how the color of these pixels are assigned to convey objects such as text, images, and user interfaces \cite{mckesson2018}. The higher the resolution of a display the more detail that can be shown on elements that are drawn. For reference, displays commonly range from standard HD at 1280x720, to very high resolutions such as 3840 x 2160, referred to as 4K. Standard HD is becoming less common while 4K is slowing gaining market share, especially on high-end monitors, TVs, and other consumer electronics.

Graphics rendering is done by defining primitives such as points, lines, and triangles in a 3D space. After these points are defined, their locations are transformed and then they are drawn on the screen through a process called rasterization \cite{mckesson2018}. Figure \ref{fig:triangle-rendering} shows a simple example of the rasterization process. The figure begins by defining a triangle primitive which is composed of three 3D vectors. The rasterizer then determines which fragments the triangle overlaps through a process called scan conversion. One way to do this is to include pixels if their centers are contained within the element. Rasterization then produces list of fragments. A fragment is an element that contributes to the final color of a pixel \cite{sellers2016}. The fragments are assigned a color in later steps of the rendering process and which may be changed when effects such as lighting are taken into account.

Although these three primitives may seem simple and insignificant they are actually the backbone of graphics rendering. Complex 3D models such as those shown in video games are actually meshes of many small, connected triangles. Squares and other quadrilaterals can be created by connecting two triangles together at their hypotenuses.

\begin{figure}[h]
	\centering
	\includegraphics[height=3cm]{triangle-rendering}
	\caption{A visualization of how a single 2D triangle is rasterized. \cite{mckesson2018}.}
	\label{fig:triangle-rendering}
\end{figure}

\subsection{Matrices}
A two-dimensional object can easily be drawn by simply defining the x and y coordinates. Three-dimensional objects will not be drawn correctly at this point because the z coordinate does not affect the size of the object. The z coordinate represents how close a point is to a viewer with larger numbers being closer, therefore as the z coordinate increases the object should increase in size as well. This can be fixed by using a projection matrix and matrix multiplication. A projection matrix effects what is visible on the screen, and how it is projected onto it \cite{sellers2016}. Each (x, y, z, w) coordinate will be multiplied by the matrix with the result being the location of the coordinate when projected onto the display.

Before the matrix can be created, the aspect ratio, field of view (FOV), z-near, and z-far variables must first be determined. The aspect ratio is the ratio between a display's width and height. z-near and z-far represent the closest and furthest possible z coordinate values respectively. The FOV is the angle of the scene that will be rendered on the screen. Figure \ref{fig:projection} illustrates of these four variables. Figure \ref{fig:matrix} shows an example of a 4x4 projection matrix used for 3D rendering. Orthogonal projection matrices exist for 2D rendering however they are outside of the scope of this paper.

@@

While a projection matrix defines how all rendered geometry appears, a model matrix can be used to transform individual elements. A model matrix is again a 4x4 matrix, and it is used to translate, rotate, and scale objects on the screen. Model matrices are very important in 3D rendering, as they allow objects to be positioned precisely in scenes without having to manually calculate vertex locations. Like the projection matrix, the model matrix is applied by multiplying the position of each vertex by the matrix. The result is the new position of the vertex with translation, rotation, and scaling applied.

\section{OpenGL}
OpenGL is both a specification and a graphics API which is commonly implemented on modern graphics cards. The specification defines how the use of API should effect the graphics card. Like other graphics APIs, this allows users to use the underlying graphics hardware portably. OpenGL's development began at a computer hardware manufacturer named Silicon Graphics Inc. (SGI). SGI created its own proprietary graphics API named IrisGL which was used on its workstations and graphics hardware. IrisGL was cleaned up and formalized into OpenGL. The first version of OpenGL was released to the public as version 1.0 in 1992 \cite{openglwiki2018}. This initial version has been revised many times, with the latest version being 4.6, which was released on July 31st, 2017 \cite{openglwiki2018}.

@@

Before OpenGL can be used, an OpenGL context (i.e\. a desktop window) must be created. Many libraries exist to do this in a cross-platform manner. One popular library that is bundled with lwjgl is the Graphics Library Framework (GLFW) which makes it very easy to create an OpenGL context with lwjgl. Figure \ref{fig:glfw-creation} shows the code necessary to initialize the GLFW framework, create an operating system window with GLFW, and then finally bind an OpenGL context to it for drawing. Now that a window and OpenGL context have been created, OpenGL methods can now be called, and the window is ready for drawing.

@@

\section{VBOs, VAOs, and Uniforms}
OpenGL provides the standard primitives used in graphics rendering --- points, lines, and triangles. All of these primitives are represented as float arrays which are passed to OpenGL for rendering. These float arrays are stored in vertex buffer objects (VBOs). VBOs are used to allocate memory on graphics hardware, and are bound along with other objects to vertex array objects (VAOs). Creating a VBO with OpenGL is a simple task, but it would be beneficial to understand how the OpenGL APIs work before going further into the subject.

@@

% TODO revise
\subsection{Indexed Rendering}
Consider the case of drawing a square. This would require two right triangles of equal dimensions, each connected at their hypotenuses. Each triangle would only have one unique coordinate, with the other two being shared at the hypotenuses. Right now this would have to be sent as six coordinates as show in Figure \ref{fig:square-coords}, instead of only four. While these two extra coordinates may seem insignificant, this inefficient use of graphics memory may cause problems, especially when rendering complex scenes with hundreds or thousands of triangles. This problem can be solved with indexed rendering.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/square.java}
        \caption{Without indexing}
        \label{fig:square-coords}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/square-indexed.java}
        \caption{With indexing.}
        \label{fig:square-indexed}
    \end{subfigure}
	\caption{Creating a square with and without indexing}
	\label{fig:square-compare}
\end{figure}

\section{The Rendering Pipeline}
Now that a VAO and VBO have been created, OpenGL is almost ready to draw. The only thing left to do is to tell OpenGL what to do with the data it receives from VAOs, VBOs, and uniforms. When OpenGL receives a draw command, it runs the VAO through a pipeline in order to determine what to draw and how to draw it. This pipeline is responsible for positioning, rasterizing, and coloring primitives. The pipeline is divided into several stages, many of which are configurable through programs written in the OpenGL shading language (GLSL). Figure \ref{fig:pipeline-flow} shows a simplified view of the OpenGL pipeline and Figure \ref{fig:pipeline-visualized} shows how the input is transformed throughout the pipeline. While OpenGL allows several of its stages to be programmed or configured by the user, some of the stages are fixed-function, meaning that OpenGL does not allow any customization to the stage. A couple of notable fixed-function stages include the vertex fetch stage, the primitive assembly stage, and per-sample processing. The vertex fetch stage is the first stage in the pipeline and feeds the pipeline with data from VAOs and VBOs. The primitive assembly stage performs an optimization known as face culling to ensure that only visible triangles are processed. Per-sample processing is the final stage in the OpenGL pipeline and ensures that previously drawn elements are not overwritten unless the new element's depth is higher through a process called depth testing, and writes fragments to the framebuffer to be displayed. % TODO revise

@@

\subsection{Textures}
Textures allows images to be stored and used while rendering. While OpenGL supports many kinds of textures, only standard 2D textures will be covered here. The process for creating a texture follows the same pattern as VBOs. A texture name must first be created and bound, and then data can be loaded into the texture. This code is show in Figure \ref{fig:texture-init}. Once the texture is initialized, it can be used in a fragment shader.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/texture.java}
        \caption{Creating a texture buffer and loading image data into it.}
        \label{fig:texture-init}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=C]{code/texture-fragment.glsl}
        \caption{Loading texture data in the fragment shader.}
        \label{fig:square-indexed}
    \end{subfigure}
	\caption{Using textures.}
	\label{fig:texture-fragment}
\end{figure}

\section{Conclusion}
With all of these concepts, a basic OpenGL program can be created. Figure \ref{fig:final-product} shows a program\footnote{Full program source available at \url{https://github.com/ShepherdJerred-homework/seminar-application}} that renders a simple textured cube that rotates. It uses all of the previously covered material and most of the code samples used above. While this program is simple, it composes many concepts and lays the foundation for more complex applications. Games can be created with only these concepts; the only further graphics work would be to render more elements and likely make them more complex.

\begin{figure}[h]
    \centering
    \includegraphics[height=8cm]{textured-cube.png}
	\caption{A spinning, textured cube.}
	\label{fig:final-product}
\end{figure}

>>>>

% Title Page
\begin{titlingpage}
    \maketitle
    \begin{abstract}
    Computer graphics is an essential component to any consumer facing computer. Efficiently rendering computer graphics requires the use of specialized hardware in the form of graphics processing units. GPUs work differently than the CPUs which programmers are experienced with. This difference is due to the GPUs approach to parallelization. Graphics APIs have been created to help programmers write performant and portable code for GPUs. This paper will introduce the core concepts of 3D graphics rendering and OpenGL, a popular and widely-supported graphics API.
    \end{abstract}
\end{titlingpage}

% Body
\section{Introduction}
Computers have become widespread and have touched many aspects of modern-day life. From new forms of entertainment such as video games to artificial intelligence, they have revolutionized the way to work, learn, and play. Computer graphics in particular have been a driving force behind the adoption of computers because they allow users to easily interact with computers. Programmers use computer graphics to display user interfaces, render video games, and even animate entire films. Every day millions of people interact with interfaces on phones, laptops, and other devices that allow them to focus on the work they are doing rather than how they communicate with their device. % TODO maybe this could still be improved

Plenty of tools to manipulate 2D and 3D graphics already exist, such as: 3D object editors, image processing applications, UI libraries, and game engines. These tools allow programmers to quickly get work done, but they may not know how their work is being accomplished. Without this understanding it may be harder for the programmer to debug or optimize their code. By learning the lower-level concepts of computer graphics, one is able to better use higher-level abstractions and tools. % TODO this is still a little awkward

This paper intends to introduce the reader to the broad concepts of graphics processing. This includes the purpose of graphics processing units, why graphics APIs exist, and an introduction to a commonly used cross-platform graphics API --- OpenGL. OpenGL will be discussed so that the reader has an overview of the concepts of OpenGL and can create an OpenGL program. This paper will focus on the use OpenGL version 3.2 which is widely supported and is very similar to the most recent version of OpenGL, which is 4.7.

\section{Background}
Computer graphics is predisposed towards parallel processing due to the repetitive and independent nature of the calculations that it requires \cite[p.~4]{sellers2016}. Graphics processing units (GPUs) were created in order to meet the unique hardware requirements of computer graphics and are fundamentally different from more traditional central processing units (CPUs). While CPUs generally have a few very fast cores, GPUs take another approach by having a large number of slow cores. Figure \ref{fig:architecture} illustrates this difference in architecture. It shows the massive difference in arithmetic logic units (ALUs) between CPUs and GPUs. These ALUs give GPUs an edge over CPUs when performing mathematical operations and are a core part of the parallel processing capability that GPUs possess.

@@

A difference also exists between the two in their approach to threading. An application running on a CPU may generally have only few threads running due to the high cost of creation and context switches. GPUs have lower cost thread creation and can more quickly context switch, and they often require tens of thousands of threads in order to fully reach their processing potential \cite{larkin2016}. Despite the slower performance of an individual GPU core, GPUs easily outperform CPUs in tasks related to graphics processing due to their parallel processing capabilities. Figure \ref{fig:performance} shows the historical theoretical performance gap between CPUs and GPUs in giga floating-point operations per second (GFLOP/s), which is a common measure of graphical computation speed. The graph shows that GPUs have a clear performance advantage, with the gap becoming exponentially larger as time goes on.

@@

Using GPUs effectively is especially important when rendering user interfaces, which generally should remain responsive at all times. Because of the architectural differences between GPUs and CPUs, GPUs are programmed differently than CPUs so that their peak performance can be reached. In addition to writing high-level code which is compiled into assembly, GPUs are also programmed by using APIs that are exposed by the manufacturer of the graphics card. GPU manufacturers may support proprietary APIs which is very common on cards that are made to be used in systems such as game consoles and arcade machines, or they may implement standardized APIs such as OpenGL, Metal, Vulkan, or DirectX.

While using a proprietary API may allow a programmer to get more performance out of the hardware they are targeting, it will severely limit the application's portability. A program that is written to run on a GPU using OpenGL should be compatible with any other GPU implementing the same version of OpenGL. These APIs help programmers to focus on the code they are writing rather than what hardware the code will be executed on. The APIs are created so that they are at a low enough level that maximum performance can be achieved while also being at a high enough level so that programmers can easily use them \cite[p.~4]{sellers2016}.

\section{Graphics Rendering}
Computer displays, such as monitors, use a 2D matrix of pixels to display images. The dimensions of this matrix are referred to as the display's resolution. Traditionally each pixel of a color display has a red, blue, and green value, which together determine the color of the pixel. Graphics rendering is how the color of these pixels are determined so that objects such as text, images, and user interfaces can be conveyed \cite{mckesson2018}. The higher the resolution of a display the more detail that can be shown on elements that are drawn. For reference, displays commonly range from standard HD at 1280x720, to very high resolutions such as 3840 x 2160, referred to as 4K. Standard HD is becoming less common while 4K is slowing gaining market share, especially on high-end monitors, TVs, and other consumer electronics.

Graphics rendering is done by positioning primitives such as points, lines, and triangles in a 3D space. After these primitives are defined, their locations are transformed, and they are drawn on the screen through a process called rasterization \cite{mckesson2018}. Figure \ref{fig:triangle-rendering} shows a simple example of the rasterization process. The figure begins by defining a triangle primitive which is composed of three 3D vectors. The rasterizer then determines which fragments the triangle overlaps through a process called scan conversion. One way to do this is to include pixels if their centers are contained within the element. Rasterization then produces list of fragments. A fragment is an element that contributes to the final color of a pixel \cite{sellers2016}. The fragments are assigned a color in later steps of the rendering process, which may be changed even further in the process when effects such as lighting are taken into account.

Although these three primitives may seem simple and insignificant, they are actually the backbone of graphics rendering. Complex 3D models such as those shown in video games are really just meshes of many small, connected triangles. Squares and other quadrilaterals can be created by connecting two triangles together at their hypotenuses.

\begin{figure}[h]
	\centering
	\includegraphics[height=3cm]{triangle-rendering}
	\caption{A visualization of how a single 2D triangle is rasterized \cite{mckesson2018}.}
	\label{fig:triangle-rendering}
\end{figure}

\subsection{Matrices}
A two-dimensional object can be drawn by simply defining its x and y coordinates. Three-dimensional objects will not be drawn correctly because the z coordinate does not affect the size of the object. The z coordinate represents how close a point is to a viewer with larger numbers being closer, therefore as the z coordinate increases the object should increase in size as well. This can be fixed by using a projection matrix and matrix multiplication. A projection matrix effects what is visible on the screen, and how it is projected onto it \cite{sellers2016}. Each (x, y, z, w) coordinate will be multiplied by a 4x4 matrix, with the result being the location of the coordinate when projected onto the display.

Before the matrix can be created, the aspect ratio, field of view (FOV), z-near, and z-far variables must be determined. The aspect ratio is equal to a display's width in pixels divided by its height in pixels. z-near and z-far represent the closest and furthest possible z coordinate values respectively. The FOV is the angle of the scene that will be rendered on the screen. Figure \ref{fig:projection} illustrates of these four variables. Figure \ref{fig:matrix} shows an example of a 4x4 projection matrix used for 3D rendering. Orthogonal projection matrices exist for 2D rendering however they are outside of the scope of this paper.

@@

While a projection matrix defines how all rendered geometry appears, a model matrix can be used to transform individual elements. Like a projection matrix, a model matrix is a 4v4 matrix. It is used to translate, rotate, and scale individual objects on the screen. Model matrices are very important in 3D rendering, as they allow objects to be positioned precisely in scenes without having to manually calculate vertex locations. Like the projection matrix, the model matrix is applied by multiplying the position of each vertex by the matrix. The result is the new position of the vertex with translation, rotation, and scaling applied.

\section{OpenGL}
OpenGL is both a specification and a graphics API which is commonly implemented on modern graphics cards. The specification defines how the use of API should affect the graphics card. Like other graphics APIs, this allows users to use the underlying graphics hardware portably. OpenGL's development began at a computer hardware manufacturer named Silicon Graphics Inc. (SGI). SGI created its own proprietary graphics API named IrisGL which was used on its workstations and graphics hardware. IrisGL was cleaned up and formalized into OpenGL. The first version of OpenGL was released to the public as version 1.0 in 1992 \cite{openglwiki2018}. This initial version has been revised many times, with the latest version being 4.6, which was released on July 31st, 2017 \cite{openglwiki2018}.

@@

Before OpenGL can be used, an OpenGL context (i.e. a desktop window) must be created. Many libraries exist to do this in a cross-platform manner. One popular library that is bundled with lwjgl is the Graphics Library Framework (GLFW), which makes it very easy to create an OpenGL context with lwjgl. Figure \ref{fig:glfw-creation} shows the code necessary to initialize the GLFW framework, create an operating system window with GLFW, and then finally bind an OpenGL context to the window for drawing. Now that a window and OpenGL context have been created, OpenGL methods can now be called, and the window is ready for drawing.

@@

\section{VBOs, VAOs, and Uniforms}
OpenGL provides the standard primitives used in graphics rendering --- points, lines, and triangles. All of these primitives are represented as float arrays which can be passed to OpenGL for rendering. These float arrays are stored in vertex buffer objects (VBOs). VBOs are used to allocate memory on graphics hardware and are bound along with other objects to vertex array objects (VAOs). Creating a VBO with OpenGL is a simple task, but it would be beneficial to understand how the OpenGL APIs work before going further into the subject.

@@

% TODO revise
\subsection{Indexed Rendering}
Consider the case of drawing a square. This would require two right triangles of equal dimensions, each connected at their hypotenuses. Each triangle would only have one unique coordinate, with the other two being shared at the hypotenuses. Right now, this would have to be sent as six coordinates as show in Figure \ref{fig:square-coords}, instead of only four. While these two extra coordinates may seem insignificant, this inefficient use of graphics memory may cause problems, especially when rendering complex scenes with hundreds or thousands of triangles. This problem can be solved with indexed rendering.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/square.java}
        \caption{Without indexing.}
        \label{fig:square-coords}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/square-indexed.java}
        \caption{With indexing.}
        \label{fig:square-indexed}
    \end{subfigure}
	\caption{Creating a square with and without indexing.}
	\label{fig:square-compare}
\end{figure}

\section{The Rendering Pipeline}
Now that a VAO and VBO have been created, OpenGL is almost ready to draw. The only thing left to do is to tell OpenGL what to do with the data it receives from VAOs, VBOs, and uniforms. When OpenGL receives a draw command, it runs the VAO through a pipeline in order to determine what to draw and how to draw it. This pipeline is responsible for positioning, rasterizing, and coloring primitives. The pipeline is divided into several stages, many of which are configurable through programs written in the OpenGL shading language (GLSL). Figure \ref{fig:pipeline-flow} shows a simplified view of the OpenGL pipeline and Figure \ref{fig:pipeline-visualized} shows how the input is transformed throughout the pipeline. While OpenGL allows several of its stages to be programmed or configured by the user, some of the stages are fixed-function, meaning that OpenGL does not allow any customization to the stage. A couple of notable fixed-function stages include the vertex fetch stage, the primitive assembly stage, and per-sample processing. The vertex fetch stage is the first stage in the pipeline and feeds the pipeline with data from VAOs and VBOs. The primitive assembly stage performs an optimization known as face culling to ensure that only visible triangles are processed. Per-sample processing is the final stage in the OpenGL pipeline and ensures that previously drawn elements are not overwritten unless the new element's depth is higher through a process called depth testing and writes fragments to the framebuffer to be displayed. % TODO revise

@@

\subsection{Textures}
Textures allow images to be stored and used while rendering. While OpenGL supports many kinds of textures, only standard 2D textures will be covered here. The process for creating a texture follows the same pattern as VBOs. A texture name must first be created and bound, and then data can be loaded into the texture. This code is show in Figure \ref{fig:texture-init}. Once the texture is initialized, it can be used in a fragment shader.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=Java]{code/texture.java}
        \caption{Creating a texture buffer and loading image data into it.}
        \label{fig:texture-init}
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
        \lstinputlisting[language=C]{code/texture-fragment.glsl}
        \caption{Loading texture data in the fragment shader.}
        \label{fig:texture-fragment}
    \end{subfigure}
	\caption{Using textures.}
	\label{fig:texture-code}
\end{figure}

\section{Conclusion}
With all of these concepts a basic OpenGL program can be created. Figure \ref{fig:final-product} shows a program\footnote{Full program source available at \url{https://github.com/ShepherdJerred-homework/seminar-application}} that renders a simple textured cube that rotates. It uses all of the previously covered material, and most of the code samples used above. While this program is simple, it composes many concepts and lays the foundation for more complex applications. Games can be created with only these concepts. The only further work with graphics would be to render more elements and make the objects more complex.

\begin{figure}[h]
    \centering
    \includegraphics[height=8cm]{textured-cube.png}
	\caption{A spinning textured cube.}
	\label{fig:final-product}
\end{figure}
